---
title: "Data Categories"
subtitle: "What the merged database supports, by fraction, covariate availability, and layering"
---

Before running analyses on the merged database, it helps to see what is actually
there. The same trace element behaves very differently depending on how the
sediment was sampled and what supporting data came with it, and not every
analysis is possible on every measurement. This page counts the measurements per
element, split by the factors that decide which analyses each subset can carry.

The counts are produced upstream in the processing pipeline and read here from
tidy CSVs, so the category definitions live in one script.

## The factors

Each measurement is classified by five factors:

1. **Fraction**: whether the reading is on the **bulk** (whole) sediment or on a
   **sieved** sub-fraction (<63 µm, the mud fraction, or <20 µm, finer still).
   A coarse sieve isolates the fine, geochemically active material, so bulk and
   sieved readings of the same element are not directly comparable and are kept
   as separate tracks.
2. **Grain size**: whether the sample carries grain-size composition, needed for
   grain-size correction and mud-fraction work.
3. **Normalisers (Fe, Al)**: whether iron and/or aluminium are available, the
   lithogenic carriers used to normalise the target metals.
4. **Organic carbon**: whether CORG/TOC is available, the organic-matter carrier.
5. **Layering**: whether the sampling event is a **single** grab or core section
   (one layer) or a **multi-layer / multi-core** event (a depth profile).

Grain-size correction and Fe/Al normalisation are only meaningful on **bulk**
samples, so factors 2 to 4 are tabulated for bulk only. Sieved fractions are
split by layering alone.

::: {.callout-note}
Covariate availability (factors 2 to 4) is read from the pipeline's
subsample-level "exist" flags. They mark that the measurand exists somewhere in
that subsample and are fraction-agnostic, so a bulk row flagged as having Fe has
Fe in the subsample but not necessarily in the bulk fraction itself. This is a
planning view (which analyses are worth attempting); each analysis script matches
the fraction exactly.
:::

```{r}
#| label: setup
library(knitr)

read_summary <- function(f) {
  d <- read.csv(file.path("data/merged_summary", f),
                stringsAsFactors = FALSE, check.names = FALSE)
  d$Category <- factor(tools::toTitleCase(d$category),
                       levels = c("Target", "Reference", "Organic"))
  d$Element  <- paste0(d$element_name, " (", d$symbol, ")")
  d
}
```

## Coverage by fraction

How many measurements each element carries in each fraction. The bulk track
dominates, but both sieved tracks carry thousands of target-metal readings, most
of them Copper and Zinc.

```{r}
#| label: coverage
cov <- read_summary("merged_coverage_fraction.csv")
kable(
  cov[order(cov$Category, cov$symbol),
      c("Category", "Element", "bulk", "sieved63", "sieved20", "other", "total")],
  col.names = c("Category", "Element", "Bulk", "Sieved <63 µm",
                "Sieved <20 µm", "Other", "Total"),
  row.names = FALSE, format.args = list(big.mark = ","))
```

The "Other" column holds a handful of readings sieved to unusual cutoffs (90 and
500 µm) and is set aside.

## Bulk sub-categories: covariate availability

Within the bulk track, how many measurements come with each covariate. This is
what decides whether an element can be grain-size corrected, Fe/Al normalised, or
related to organic carbon.

```{r}
#| label: bulk-factors
bf <- read_summary("merged_bulk_factors.csv")
kable(
  bf[order(bf$Category, bf$symbol),
     c("Category", "Element", "bulk", "grain", "fe_or_al", "org")],
  col.names = c("Category", "Element", "Bulk total", "With grain size",
                "With Fe or Al", "With organic C"),
  row.names = FALSE, format.args = list(big.mark = ","))
```

Coverage is uneven. Copper and Zinc have the most bulk readings but the smallest
proportion with normalisers (roughly a quarter), because much of their data comes
from monitoring programmes that did not measure Fe/Al. The less common metals
(Cobalt, Manganese, Molybdenum, Selenium) are better supported, often measured on
full geochemical suites. Organic carbon travels with Copper and Zinc far more
often than the normalisers do.

## Layering: single vs multi-layer / core

Whether each measurement comes from a single grab or core section, or from a
multi-layer / multi-core event (a depth profile). Multi-layer events open up
downcore analyses; single events are the bulk of the surface-sediment data.

```{r}
#| label: layering
lay <- read_summary("merged_layering.csv")
kable(
  lay[order(lay$Category, lay$symbol),
      c("Category", "Element",
        "bulk_single", "bulk_multi",
        "sieved63_single", "sieved63_multi",
        "sieved20_single", "sieved20_multi")],
  col.names = c("Category", "Element",
                "Bulk single", "Bulk multi",
                "<63 single", "<63 multi",
                "<20 single", "<20 multi"),
  row.names = FALSE, format.args = list(big.mark = ","))
```

Most data is single-layer surface sediment. Multi-layer events are a meaningful
minority of the bulk track (roughly a fifth for Copper and Zinc, more for Cobalt,
Molybdenum and Selenium) but are rare in the sieved tracks, so downcore work is
essentially a bulk-only option.

## Suggested analytical approaches

Reading the three tables together, the merged data splits into a few workable
groups, each suiting a different analysis:

- **Bulk, with grain size**: the largest analysable group. Supports the
  **grain-size** analysis (mud fraction vs metal) and grain-size correction, the
  first-line control on the target metals. Available for every target element,
  strongest for Cobalt, Manganese and Selenium.
- **Bulk, with Fe or Al**: supports **Fe/Al normalisation** (metal vs lithogenic
  carrier), the standard geochemical baseline. Well covered for the full-suite
  metals; for Copper and Zinc it is a large absolute count but a minority of their
  readings, so normalised and un-normalised subsets should be reported separately.
- **Bulk, with organic carbon**: supports the **organic-carbon** analysis (metal
  vs CORG/TOC), the organic-matter carrier. Broadest covariate for Copper and
  Zinc, so the natural control where normalisers are scarce.
- **Bulk, multi-layer / core**: the subset for **downcore / depth-profile**
  analyses (metal vs sediment depth). Bulk only in practice, best for Copper,
  Zinc, Cobalt and Selenium.
- **Sieved (<63 µm and <20 µm)**: supports a **bulk-versus-sieved comparison**
  (how much the fine fraction concentrates each metal) and analyses run within a
  single fraction. Mostly Copper, Zinc and Manganese; too sparse for Iodine and
  Selenium.
- **All fractions**: the spatial (**depth and distance to coast**) and temporal
  (**sampling year**) settings apply to every measurement and are the context
  layer for all of the above.

The clean-database site runs the grain-size, Fe/Al, organic-carbon, spatial and
temporal analyses per source; the same analyses can be repeated here on the merged
database, now with the fraction and layering splits made explicit above.
