---
title: "Depth Profiles"
subtitle: "How the target metals change downcore, within multi-layer sediment cores"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Some sampling events are **cores**, sliced into several depth layers rather than a
single surface grab. Reading a metal down through a core separates recent input from
the natural background beneath it: a metal that concentrates in the top slices and
falls with depth is **surface-enriched**, the fingerprint of recent (often
anthropogenic) deposition, while a flat or rising profile points to a stable or
buried source, or to diagenetic remobilisation.

This analysis has no per-source clean counterpart: it needs the merged database's
multi-layer events (`n_layers > 1`). Because the absolute concentration differs
enormously between sites, profiles are compared in **relative** terms, both measures
cancelling a site's baseline level:

- the **within-core trend**, a rank correlation (Spearman rho) of concentration
  against layer midpoint depth (negative = falls with depth = surface-enriched);
- each measurement divided by **its own core's median**, pooled across cores into an
  average downcore shape.

Bulk samples only; distributional outliers dropped; depth is the subsample interval
midpoint.

```{r}
#| label: setup
library(dplyr)
library(ggplot2)
library(knitr)

ddir <- "data/merged_summary"
rd <- function(f) read.csv(file.path(ddir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
trends <- rd("merged_depthprofile_trends.csv")
enrich <- rd("merged_depthprofile_enrichment.csv")
pooled <- rd("merged_depthprofile_pooled.csv")
crho   <- rd("merged_depthprofile_core_rho.csv")

elem_levels <- c("CO", "CU", "MN", "MO", "SE", "ZN")
bin_levels  <- c("0-2", "2-5", "5-10", "10-20", "20-40", ">40")
lev <- elem_levels[elem_levels %in% unique(trends$symbol)]
theme_set(theme_minimal(base_size = 12))
```

## 1. Within-core trend

The Spearman correlation of concentration with depth, computed per core (cores with
at least three layers), then summarised per element. A negative median rho, and a
high share of cores decreasing with depth, mark surface enrichment.

```{r}
#| label: trend-table
trends |>
  mutate(Element = factor(symbol, levels = lev)) |>
  arrange(Element) |>
  select(Element, n_cores, median_rho, pct_decrease, pct_increase, pct_flat) |>
  kable(col.names = c("Element", "Cores", "Median rho", "% decreasing",
                      "% increasing", "% flat"),
        format.args = list(big.mark = ","))
```

**Manganese** is the clearest case: its median rho is strongly negative and most of
its cores fall with depth. This is diagenetic, manganese oxides precipitate at the
oxic surface and dissolve below, concentrating Mn near the top. **Copper** and
**Zinc** are more moderately surface-enriched (about half their cores decrease), the
pattern expected from recent deposition in a subset of contaminated cores.
**Cobalt** and **Molybdenum** lean the other way, richer at depth: molybdenum in
particular accumulates in the reducing sediment below the surface, so it reads as a
redox proxy rather than a recent-input tracer. **Selenium** shows no downcore
structure.

```{r}
#| label: rho-fig
#| fig-cap: "Distribution of within-core trend (Spearman rho) per element. Below the dashed line = surface-enriched cores."
#| fig-width: 7.5
#| fig-height: 4
crho |>
  mutate(Element = factor(symbol, levels = lev)) |>
  ggplot(aes(Element, rho)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_boxplot(fill = "#4477AA", alpha = 0.6, outlier.size = 0.5,
               outlier.alpha = 0.3) +
  labs(x = NULL, y = "within-core rho (concentration vs depth)")
```

## 2. Surface versus deep enrichment

A second, simpler view: the ratio of the shallowest layer's concentration to the
deepest layer's, per core (cores with at least two layers). Above 1 means
surface-enriched. Manganese stands out again (median 1.7, four cores in five higher
at the surface); copper and zinc are enriched in just over half their cores.

```{r}
#| label: enrich-table
enrich |>
  mutate(Element = factor(symbol, levels = lev)) |>
  arrange(Element) |>
  select(Element, n_cores, median_ratio, p25, p75, pct_surf_higher) |>
  kable(col.names = c("Element", "Cores", "Median surf/deep", "P25", "P75",
                      "% surface higher"),
        format.args = list(big.mark = ","))
```

## 3. Average downcore shape

Each measurement relative to its own core's median, pooled by depth bin. Pooling
levels flattens the median profile (most cores are near-uniform, and the enriched
minority is diluted), so the curves stay close to 1; the interquartile ribbon shows
where the enriched cores pull the spread. Manganese carries the widest near-surface
spread, consistent with its strong but variable surface enrichment.

```{r}
#| label: pooled-fig
#| fig-cap: "Relative concentration (value / core median) against depth, pooled across cores. Line = median, ribbon = interquartile range. Depth axis inverted."
#| fig-width: 8.5
#| fig-height: 5
pooled |>
  mutate(Element = factor(symbol, levels = lev)) |>
  ggplot(aes(bin_center, median_rel)) +
  geom_ribbon(aes(ymin = p25, ymax = p75), fill = "#4477AA", alpha = 0.2) +
  geom_line(colour = "#4477AA") +
  geom_point(size = 1, colour = "#4477AA") +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey60") +
  scale_x_reverse(breaks = c(1, 3.5, 7.5, 15, 30, 55), labels = bin_levels) +
  coord_flip() +
  facet_wrap(~ Element, scales = "free_x") +
  labs(x = "depth (cm, midpoint)", y = "relative concentration (value / core median)")
```

::: {.callout-note}
Surface enrichment here is a population summary, not a per-site verdict: an
individual core can be enriched, buried or flat regardless of its element's average.
The trend is strongest and most consistent for manganese (diagenetic), moderate and
input-driven for copper and zinc, and reversed for molybdenum (redox accumulation at
depth). Iodine has too few multi-layer cores to report.
:::
