---
title: "Bulk vs Sieved"
subtitle: "How much sieving to the fine fraction concentrates each metal"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Sieving sediment to a fine cutoff (<63 µm mud, or <20 µm finer still) discards the
coarse, metal-poor grains, so a sieved analysis reads higher than a bulk (whole
sample) one for the same metal. The EFSA specification warns that the two are not
directly comparable; this page quantifies the **enrichment factor** (sieved / bulk)
so the tracks can be related.

The naive way to measure it, comparing the pooled median sieved value to the pooled
median bulk value, is **confounded by location**: sieved samples tend to come from
muddier, often more contaminated places, so a pooled ratio blends the sieving effect
with a site-selection effect. The stronger measure is **paired within site**: at a
site that carries both a bulk and a sieved value for the same element, their ratio is
a direct enrichment factor with location held fixed. This page leads with the paired
result and shows the pooled one alongside, because the gap between them is itself the
finding.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

edir <- "data/merged_summary"
rd <- function(f) read.csv(file.path(edir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
paired <- rd("merged_enrichment_paired.csv")
pairs  <- rd("merged_enrichment_pairs.csv")
pooled <- rd("merged_enrichment_pooled.csv")

elem_levels  <- c("CO", "CU", "MN", "MO", "SE", "ZN")
cut_levels   <- c("<63um", "<20um")
lev <- elem_levels[elem_levels %in% unique(paired$symbol)]
theme_set(theme_minimal(base_size = 12))
```

## 1. Within-site enrichment factor

At sites carrying both fractions of the same element, the ratio of the sieved value
to the bulk value (at least 15 paired sites). Ratios are skewed, so summarised by
median and interquartile range. Enrichment is **modest**: medians sit around 1.0 to
1.4, and the finer <20 µm cutoff concentrates more than the <63 µm one, as expected.

```{r}
#| label: paired-table
paired |>
  mutate(Element = factor(symbol, levels = lev),
         cutoff = factor(cutoff, levels = cut_levels)) |>
  arrange(Element, cutoff) |>
  select(Element, cutoff, n_sites, median_ratio, p25, p75, geomean, pct_enriched) |>
  kable(col.names = c("Element", "Sieve", "Paired sites", "Median", "P25", "P75",
                      "Geomean", "% enriched"),
        format.args = list(big.mark = ","))
```

```{r}
#| label: paired-fig
#| fig-cap: "Within-site enrichment factor (sieved / bulk) per element and cutoff (log scale). Dashed line = no enrichment (ratio 1)."
#| fig-width: 8
#| fig-height: 4
pairs |>
  mutate(Element = factor(symbol, levels = lev),
         cutoff = factor(cutoff, levels = cut_levels)) |>
  ggplot(aes(Element, ratio, fill = cutoff)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_boxplot(alpha = 0.6, outlier.size = 0.4, outlier.alpha = 0.25) +
  scale_y_log10() +
  scale_fill_manual(values = c("<63um" = "#CCBB44", "<20um" = "#228833")) +
  labs(x = NULL, y = "enrichment factor (sieved / bulk, log)", fill = "sieve")
```

## 2. Paired versus pooled

The pooled median-to-median ratio (all bulk and all sieved data, not matched by site)
tells a much stronger story than the paired one, and that difference is the location
confounding at work. For manganese and zinc the pooled <20 µm ratio is three to four
times, but the honest within-site factor is under 1.5: most of the apparent
"enrichment" is that sieved samples were taken in finer, richer sediment to begin
with.

```{r}
#| label: compare-table
pool_long <- pooled |>
  transmute(symbol,
            `<63um` = ratio_s63,
            `<20um` = ratio_s20) |>
  pivot_longer(c(`<63um`, `<20um`), names_to = "cutoff", values_to = "pooled_ratio")

paired |>
  select(symbol, cutoff, median_ratio) |>
  inner_join(pool_long, by = c("symbol", "cutoff")) |>
  mutate(Element = factor(symbol, levels = lev),
         cutoff = factor(cutoff, levels = cut_levels)) |>
  arrange(Element, cutoff) |>
  transmute(Element, Sieve = cutoff,
            `Within-site` = median_ratio,
            `Pooled` = pooled_ratio,
            `Pooled / paired` = round(pooled_ratio / median_ratio, 1)) |>
  kable(format.args = list(big.mark = ","))
```

::: {.callout-note}
The practical message: a bulk-to-sieved conversion built on pooled medians would
over-correct badly. The within-site factors here (roughly 1.1 to 1.5, larger for the
finer cutoff) are the defensible starting point, though they still carry a wide
spread from site to site, so bulk and sieved are best kept as separate tracks and
converted only with care. Molybdenum and selenium have too few paired sites to report
and are set aside; iodine has none.
:::
