---
title: "Geochemical Facies"
subtitle: "Unsupervised multi-element clusters in the sediment chemistry, and where they fall"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The other analyses ask directed questions (does enrichment track depth, year, grain
size). This one asks an open one: left to itself, does the sediment chemistry fall
into distinct multi-element **signatures**, or facies? We cluster the samples in
**chemistry space** with K-means and then look at where the clusters land on the map,
so the grouping is driven by composition, not by location.

The features are the four best-covered targets, **copper, zinc, manganese and
cobalt** (the wider seven-target set has no samples with all elements measured). Each
sample must carry all four. Bulk and sieved are clustered **separately** (they are
never pooled), and each fraction is treated on the basis that suits it:

- **Bulk** clusters on each metal as a **ratio to iron** (metal / Fe). Normalising to
  the grain-size carrier means a cluster reflects an **enrichment signature** rather
  than simply how muddy the grab was (raw metals would mostly separate mud from sand).
- **Sieved** clusters on the **raw** value: the sieve cut already fixes the grain
  size, so no iron normalisation is applied.

Every feature is log-transformed (metals are log-normal) and standardised, so no
single element dominates the distance. Distributional outliers are dropped.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

cdir <- "data/merged_summary"
rd <- function(f) read.csv(file.path(cdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
ksel <- rd("merged_clustering_kselect.csv")
cent <- rd("merged_clustering_centroids.csv")
asg  <- rd("merged_clustering_assignments.csv")
meta <- rd("merged_clustering_meta.csv")

CORE <- c("CU", "ZN", "MN", "CO")
PAL  <- c("#4477AA", "#EE6677", "#228833", "#CCBB44")  # Paul Tol, colourblind-safe
theme_set(theme_minimal(base_size = 12))
```

The cohort is **`r format(meta$n_bulk, big.mark = ",")`** bulk and
**`r format(meta$n_sieved, big.mark = ",")`** sieved samples with all four elements.

## Choosing the number of clusters

The number of facies is not fixed in advance. We sweep k from 2 to 8, and for each
score how cleanly separated the clusters are with the **mean silhouette width**
(higher is better). The chosen k is the peak, taken per fraction.

```{r}
#| label: k-fig
#| fig-cap: "Mean silhouette width against the number of clusters. Bulk peaks at four facies, sieved at two."
#| fig-width: 7
#| fig-height: 3
kbest <- ksel |> group_by(fraction) |> slice_max(mean_sil, n = 1)
ksel |>
  ggplot(aes(k, mean_sil, colour = fraction)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  geom_point(data = kbest, size = 3.4, shape = 21, stroke = 1.1, fill = "white") +
  scale_colour_manual(values = c(bulk = "#4477AA", sieved = "#EE6677"), name = NULL) +
  scale_x_continuous(breaks = 2:8) +
  labs(x = "number of clusters (k)", y = "mean silhouette width")
```

Bulk settles on **four** facies (the silhouette is nearly flat from three to five, and
four adds a distinct, interpretable group), sieved on a clean **two**. The thinner,
more uniform sieved set simply does not support finer structure.

## The facies signatures

Each facies is summarised by its median level of the four elements. To read them on
one scale, the heatmap shows each cluster's median as a **fold-change from the overall
median** of that element and fraction (log2): red is enriched, blue depleted. The raw
per-cluster medians (metal / Fe for bulk, mg/kg for sieved) are in the table below.

```{r}
#| label: sig-fig
#| fig-cap: "Facies signatures as log2 fold-change from the overall median (red enriched, blue depleted). Cluster numbers run low to high overall metal within each fraction."
#| fig-width: 7.5
#| fig-height: 3.6
sig <- asg |>
  pivot_longer(all_of(paste0("feat_", CORE)),
               names_to = "element", values_to = "v") |>
  mutate(element = factor(sub("^feat_", "", element), levels = CORE)) |>
  group_by(frac_class, element) |>
  mutate(overall = median(v)) |>
  group_by(frac_class, cluster, element) |>
  summarise(fold = log2(median(v) / overall[1]), .groups = "drop")

sig |>
  ggplot(aes(element, factor(cluster), fill = fold)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%+.1f", fold)), size = 3) +
  scale_fill_gradient2(low = "#4477AA", mid = "grey95", high = "#EE6677",
                       midpoint = 0, name = "log2 fold") +
  facet_wrap(~ frac_class, scales = "free_y") +
  labs(x = NULL, y = "cluster") +
  theme(panel.grid = element_blank())
```

```{r}
#| label: sig-table
cent |>
  transmute(Fraction = frac_class, Cluster = cluster, n,
            Element = factor(feature, levels = CORE), median) |>
  pivot_wider(names_from = Element, values_from = median) |>
  arrange(Fraction, Cluster) |>
  kable(caption = "Per-facies median level (bulk: metal / Fe; sieved: mg/kg).")
```

Reading the bulk facies (metal relative to iron):

- **Cluster 1** is depleted in copper, zinc and cobalt (manganese near typical), a
  low-enrichment, iron-rich background.
- **Cluster 2** is the large baseline group (over half the samples), close to typical
  on every element.
- **Cluster 3** stands out on **manganese** (Mn / Fe roughly three times baseline)
  with the others near typical, the signature of manganese-oxide rich, often
  more oxic sediment, a redox facies rather than a contamination one.
- **Cluster 4** is enriched in **copper and zinc** together (about double) while
  cobalt and manganese stay ordinary, the classic trace-metal contamination
  fingerprint.

The two sieved facies are simpler: a lower-metal group and a higher-metal group
(cluster 2 carries several times the manganese and zinc), more a gradient of overall
load than separate chemical types.

## Where the facies fall

Clustering used no location, so the geography is an independent check. If the facies
were noise they would scatter evenly; instead they concentrate.

```{r}
#| label: map-bulk
#| fig-cap: "Bulk facies on the map. The copper-zinc enriched facies (cluster 4) clusters in the industrialised North Sea and coastal Norway; the manganese facies (cluster 3) leans to deeper, more oxic basins."
#| fig-width: 8
#| fig-height: 5.6
map_facies <- function(fr, cols) {
  d <- asg |> filter(frac_class == fr) |> arrange(cluster)
  ggplot(d, aes(longitude, latitude, colour = factor(cluster))) +
    borders("world", fill = "grey93", colour = "grey80") +
    geom_point(size = 0.9, alpha = 0.7) +
    coord_quickmap(xlim = c(-20, 40), ylim = c(36, 82)) +
    scale_colour_manual(values = cols, name = "cluster") +
    guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
    labs(x = NULL, y = NULL) +
    theme(panel.grid = element_line(colour = "grey93"))
}
map_facies("bulk", PAL)
```

```{r}
#| label: map-sieved
#| fig-cap: "Sieved facies on the map. Cluster numbering is per fraction and does not carry over from the bulk map."
#| fig-width: 8
#| fig-height: 5.6
map_facies("sieved", PAL[c(1, 2)])
```

The pattern is not random: the copper-zinc facies picks out the North Sea and the
Norwegian coast where the pressure is, and the manganese facies favours the deeper
open basins, exactly where composition, not chance, would put them.

::: {.callout-note}
These are groupings by chemistry, shown in space, not spatial regions: the map is a
check on the clustering, not its input. The cohort is limited to samples measuring all
four of copper, zinc, manganese and cobalt, which skews it toward the better
characterised programmes and regions, so absent areas mean unmeasured, not
uncontaminated. The number of facies is data-driven (the silhouette peak), and a
different, defensible k would split or merge the softer groups; the copper-zinc and
manganese bulk signatures are the robust, interpretable ones. Molybdenum, selenium and
iodine are too sparsely co-measured to enter the feature set.
:::
