---
title: "Geochemical Regions"
subtitle: "Clustering on location and composition together, to map contiguous chemical regions"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The first two Clustering pages take opposite halves of the problem:
[Geochemical Facies](./geochemical-facies.qmd) groups by **chemistry alone** (a facies
can be scattered across the map), and [Spatial Hotspots](./spatial-hotspots.qmd) groups
by **location** around a single elevated element. This page uses **both at once**:
cluster the samples on their position **and** their full composition together, so a
cluster is a spatially-contiguous region that shares a geochemical signature, a
regionalisation of the sea floor by chemistry.

The awkward question is how much **geography** should count against **chemistry**. We
make it explicit rather than hidden. Every feature is standardised; then each block
(2 location dimensions, 4 chemistry dimensions) is scaled to equal total weight, and a
single **location weight** multiplies the location block: zero is pure chemistry (the
facies page), a large value is pure geography (a plain map partition), and one balances
them. To compare cleanly with the facies page, the number of clusters is held at the
facies value (four for bulk, two for sieved) and the **location weight is swept**.

The features follow the facies rule: copper, zinc, manganese and cobalt, **bulk** on
the iron ratio, **sieved** on the raw value, log-transformed and standardised. Bulk and
sieved are clustered separately, complete cases only, outliers dropped.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

rdir <- "data/merged_summary"
rd <- function(f) read.csv(file.path(rdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
sweep <- rd("merged_regions_weightsweep.csv")
asg   <- rd("merged_regions_assignments.csv")
sig   <- rd("merged_regions_signature.csv")
meta  <- rd("merged_regions_meta.csv")

CORE <- c("CU", "ZN", "MN", "CO")
PAL  <- c("#4477AA", "#EE6677", "#228833", "#CCBB44")
theme_set(theme_minimal(base_size = 12))
```

The cohort is the facies cohort: **`r format(meta$n_bulk, big.mark = ",")`** bulk and
**`r format(meta$n_sieved, big.mark = ",")`** sieved samples with all four elements.

## The weighting tradeoff

As location weighs more, clusters get geographically tighter (lower geographic spread)
but chemically less pure (higher chemical spread). Each curve is scaled to its
chemistry-only value so the two read on one axis.

```{r}
#| label: sweep-fig
#| fig-cap: "Within-cluster geographic and chemical spread against the location weight, each relative to the chemistry-only value. For bulk, a little geography (around 0.5) tightens the map while chemistry stays as pure or purer; past that it is a tradeoff. For sieved the curves barely move: the sieved data is geographically confined, so location adds almost nothing."
#| fig-width: 8
#| fig-height: 3.4
sweep |>
  pivot_longer(c(geo_km, chem), names_to = "metric", values_to = "v") |>
  group_by(fraction, metric) |>
  mutate(rel = v / v[loc_weight == 0]) |>
  ungroup() |>
  mutate(metric = recode(metric, geo_km = "geographic spread",
                         chem = "chemical spread")) |>
  ggplot(aes(loc_weight, rel, colour = metric)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.6) +
  geom_vline(xintercept = meta$loc_weight_main, linetype = "dashed",
             colour = "grey60") +
  scale_colour_manual(values = c("geographic spread" = "#4477AA",
                                 "chemical spread" = "#EE6677"), name = NULL) +
  facet_wrap(~ fraction) +
  labs(x = "location weight", y = "spread (relative to chemistry-only)")
```

The dashed line is the balanced weight (one) used for the maps below.

## The regions at a balanced weight

With location and chemistry balanced, the bulk samples fall into four contiguous
regions, and the facies signatures from the first page reappear **attached to places**.

```{r}
#| label: sig-fig
#| fig-cap: "Region signatures as log2 fold-change from the overall median (red enriched, blue depleted), bulk regions. The same copper-zinc and manganese contrasts as the facies page, now tied to geography."
#| fig-width: 7
#| fig-height: 2.8
sig |>
  filter(frac_class == "bulk") |>
  mutate(feature = factor(feature, levels = CORE)) |>
  group_by(feature) |>
  mutate(fold = log2(median / median(median))) |>
  ungroup() |>
  ggplot(aes(feature, factor(region), fill = fold)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%+.1f", fold)), size = 3) +
  scale_fill_gradient2(low = "#4477AA", mid = "grey95", high = "#EE6677",
                       midpoint = 0, name = "log2 fold") +
  labs(x = NULL, y = "region") +
  theme(panel.grid = element_blank())
```

```{r}
#| label: region-table
sig |>
  distinct(frac_class, region, n, sea_area, lon, lat) |>
  transmute(Fraction = frac_class, Region = region, n,
            `main sea area` = sea_area, lon, lat) |>
  arrange(Fraction, Region) |>
  kable(caption = "Regions at the balanced weight: size, dominant sea area and centroid.")
```

The bulk regions read as a chemistry-aware map of the shelf:

- **Region 3, the Norwegian Sea** is manganese-rich (Mn / Fe well above typical), the
  deep-basin, oxic manganese signature the facies page found, here pinned to the
  Norwegian Sea.
- **Region 4, the southern North Sea** is the copper-zinc region (zinc almost double
  typical), the contamination signature sitting where the pressure is.
- **Region 1 (North Sea)** is a low-copper background and **Region 2 (Barents Sea)** is
  manganese-poor: the quieter northern and background areas.

The **sieved** samples, by contrast, split into two regions that are **both in the
North Sea** and differ by chemistry, not place (a lower-metal and a several-times
higher-metal group). That is the weighting curve made concrete: because the sieved data
sits almost entirely in one area, geography has nothing to separate, so the combined
clustering falls back to the facies answer.

```{r}
#| label: region-map
#| fig-cap: "Regions at the balanced weight. Bulk resolves into contiguous chemical regions (Norwegian Sea, Barents, and two North Sea regions); sieved stays within the North Sea. Region numbering is per fraction."
#| fig-width: 8.5
#| fig-height: 5.6
world <- ggplot2::map_data("world")
d <- asg |> arrange(region)

ggplot() +
  geom_polygon(data = world, aes(long, lat, group = group),
               fill = "grey93", colour = "grey80", linewidth = 0.2,
               inherit.aes = FALSE) +
  geom_point(data = d, aes(longitude, latitude, colour = factor(region)),
             size = 0.7, alpha = 0.7) +
  coord_quickmap(xlim = c(-20, 40), ylim = c(36, 82)) +
  scale_colour_manual(values = PAL, name = "region") +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  facet_wrap(~ frac_class) +
  labs(x = NULL, y = NULL) +
  theme(panel.grid = element_line(colour = "grey93"))
```

::: {.callout-note}
The location weight is a **choice, not a truth**: it sets how much geography is allowed
to shape the grouping, and the map would look tidier (more geographic) or more scattered
(more chemical) at other settings. The balanced weight is shown because it is the honest
midpoint, and the weighting curve is given so the choice is transparent. A useful
by-product is the bulk sweet spot around a weight of one-half, where a little geography
tightens the regions while chemistry stays as pure: the facies were spatially coherent
enough that location resolves borderline samples rather than fighting the chemistry. As
on the other Clustering pages the cohort is limited to samples with all four of copper,
zinc, manganese and cobalt, so coverage, not chemistry, decides which areas appear.
:::
