---
title: "Grain Size"
subtitle: "The bulk grain-size control on the target metals, pooled across the merged database"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Trace-element concentrations in sediment depend strongly on grain size: finer
material concentrates metals, so a **sieved <63 µm** analysis is not comparable to
a **bulk** (whole-sample) one. This page asks the same question as the per-source
grain-size analysis on the clean databases, but on the single merged database,
where cross-source duplicates are already removed. So it **pools** the four sources
that carry grain size (Mareano, Vannmiljø, ICES-DOME, MUDAB; 4Demon has none) into
one best estimate per element, rather than reporting each apart.

Two things the pooled, merged data adds over the per-source view:

- **Outliers are dropped.** The distributional `outlier_flag` (mostly registration
  errors, see [Outlier Flagging](./outlier-flagging.qmd)) is excluded from the
  concentration-versus-fines fits, so a few impossible values do not swing the
  slopes.
- **One relationship per element.** Deduplication means a reading shared by two
  sources is counted once, so the pooled fit is not double-weighted.

All depths are kept but reported by the EFSA depth bands (0 to 5, 5 to 40, over
40 cm), assigned from the subsample interval midpoint.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

gdir <- "data/merged_summary"
rd <- function(f) read.csv(file.path(gdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
frac  <- rd("merged_grainsize_fraction_summary.csv")
fines <- rd("merged_grainsize_fines_summary.csv")
concf <- rd("merged_grainsize_conc_vs_fines.csv")
bvs   <- rd("merged_grainsize_bulk_vs_sieved.csv")
pts   <- rd("merged_grainsize_targets_fines.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
band_levels <- c("0-5", "5-40", ">40", "unknown")
src_levels  <- c("All (pooled)", "Mareano", "Vannmiljø", "ICES-DOME", "MUDAB")
frac_lab    <- c(bulk = "Bulk", sieved63 = "Sieved <63 µm",
                 sieved20 = "Sieved <20 µm", sieved_other = "Sieved (other)")
theme_set(theme_minimal(base_size = 12))
```

## 1. Sample fraction: bulk vs sieved

How the target measurements divide between bulk (whole sample) and the two sieved
tracks (<63 µm and <20 µm). The bulk track dominates, but both sieved tracks carry
thousands of readings, mostly Copper, Zinc and Manganese.

```{r}
#| label: fraction-table
bvs |>
  mutate(fraction = frac_lab[fraction],
         Element = factor(symbol, levels = elem_levels)) |>
  select(Element, fraction, n) |>
  pivot_wider(names_from = fraction, values_from = n, values_fill = 0) |>
  arrange(Element) |>
  kable(format.args = list(big.mark = ","))
```

## 2. The <63 µm fines

The percentage of material finer than 63 µm (the clay + silt mud), by depth band,
pooled across the grain-size sources. `mud_dominated` counts subsamples that are at
least 50 % fines. Deeper layers are muddier: the fines median climbs from about
37 % near the surface to 64 % below 40 cm.

```{r}
#| label: fines-table
fines |>
  mutate(band = factor(band, levels = band_levels)) |>
  arrange(band) |>
  select(band, n_subsamples, n_with_fines, mean, median, p25, p75, mud_dominated) |>
  kable(col.names = c("Depth band (cm)", "Subsamples", "With fines",
                      "Mean %", "Median %", "P25", "P75", "Mud-dominated"),
        format.args = list(big.mark = ","))
```

## 3. Bulk versus sieved concentration

Sieving to the fine fraction lifts the concentration well above bulk for the
grain-size-controlled metals. Pooled medians: the <20 µm fraction runs roughly two
to four times the bulk level. Molybdenum is the exception (redox / organic
controlled, not grain-size controlled), so its sieved values sit at or below bulk.

```{r}
#| label: bvs-table
bvs |>
  filter(fraction %in% c("bulk", "sieved63", "sieved20")) |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  select(Element, fraction, median) |>
  pivot_wider(names_from = fraction, values_from = median) |>
  transmute(Element,
            `Bulk` = bulk,
            `<63 µm` = sieved63,
            `<20 µm` = sieved20,
            `<63 / bulk` = round(sieved63 / bulk, 1),
            `<20 / bulk` = round(sieved20 / bulk, 1)) |>
  arrange(Element) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: bvs-fig
#| fig-cap: "Median target concentration by fraction (mg/kg, log scale), pooled and outlier-free."
#| fig-width: 7.5
#| fig-height: 3.6
bvs |>
  filter(fraction %in% c("bulk", "sieved63", "sieved20"),
         symbol %in% c("CO", "CU", "MN", "MO", "ZN")) |>
  mutate(Element = factor(symbol, levels = elem_levels),
         fraction = factor(frac_lab[fraction],
                           levels = c("Bulk", "Sieved <63 µm", "Sieved <20 µm"))) |>
  ggplot(aes(Element, median, fill = fraction)) +
  geom_col(position = position_dodge()) +
  scale_y_log10() +
  scale_fill_manual(values = c("Bulk" = "#4477AA", "Sieved <63 µm" = "#CCBB44",
                               "Sieved <20 µm" = "#228833")) +
  labs(x = NULL, y = "median concentration (mg/kg, log)", fill = NULL)
```

## 4. The grain-size effect (bulk samples)

Within **bulk** samples, where the reported value describes the whole sample,
concentration rises with the fines content, as expected. The pooled Spearman
correlation of concentration with `fines_lt63` is clearly positive for the
lithogenic metals (Cobalt, Manganese, Copper, Zinc) and near zero for Molybdenum
and Selenium, which track organic matter and redox rather than grain size. Iodine
carries no bulk grain-size data and drops out.

```{r}
#| label: concf-pooled
concf |>
  filter(Source == "All (pooled)", band %in% c("0-5", "5-40", ">40")) |>
  mutate(Element = factor(symbol, levels = elem_levels),
         band = factor(band, levels = band_levels)) |>
  select(Element, band, n, rho) |>
  pivot_wider(names_from = band, values_from = c(n, rho), values_fill = NA) |>
  transmute(Element,
            `n (0-5)` = `n_0-5`,   `rho (0-5)` = `rho_0-5`,
            `n (5-40)` = `n_5-40`, `rho (5-40)` = `rho_5-40`,
            `n (>40)` = `n_>40`,   `rho (>40)` = `rho_>40`) |>
  arrange(Element) |>
  kable(format.args = list(big.mark = ","))
```

Broken out by source (all depths together), the association is tightest in the
Mareano data, whose bulk samples and clean grain-size give the clearest signal.

```{r}
#| label: concf-source
concf |>
  filter(Source != "All (pooled)") |>
  mutate(Source = factor(Source, levels = src_levels),
         Element = factor(symbol, levels = elem_levels)) |>
  select(Source, Element, n, rho) |>
  pivot_wider(names_from = Element, values_from = rho, id_cols = Source,
              names_expand = TRUE) |>
  select(where(~ !all(is.na(.)))) |>
  arrange(Source) |>
  kable(caption = "Spearman rho, concentration vs % fines (bulk, all depths).")
```

```{r}
#| label: concf-fig
#| fig-cap: "Bulk target concentration against the <63 µm fines content, pooled across sources (log y, linear fit). Outliers excluded."
#| fig-width: 8
#| fig-height: 5.5
pts |>
  filter(is.na(outlier_flag) | outlier_flag == "",
         symbol %in% c("CO", "CU", "MN", "ZN"),
         !is.na(fines_lt63), value_std > 0) |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  ggplot(aes(fines_lt63, value_std)) +
  geom_point(alpha = 0.08, size = 0.5, colour = "#4477AA") +
  geom_smooth(method = "lm", formula = y ~ x, colour = "#AA3377", se = FALSE) +
  scale_y_log10() +
  facet_wrap(~ Element, scales = "free_y") +
  labs(x = "fines <63 µm (%)", y = "concentration (mg/kg, log)")
```

::: {.callout-note}
The grain-size effect is measured on bulk samples only: a sieved <63 µm value is of
the fine fraction, not the whole sample, so relating it to the whole-sample fines
would not be meaningful. Molybdenum, Selenium and Iodine are kept in the tables but
are noisier (lower counts, and not purely grain-size controlled), so grain-size
correction is the wrong control for them.
:::
