---
title: "Fe/Al Normalisation"
subtitle: "Normalising the target metals to iron and aluminium, pooled across the merged database"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Finer, more aluminosilicate-rich sediment carries more of both the lithogenic
normalisers (iron, aluminium) and the trace metals, so dividing a metal by Fe (or
Al) corrects that grain-size and mineralogical variation. A strong metal to
normaliser relationship means the metal is naturally (lithogenically) controlled,
the baseline an enrichment-factor or pristine determination rests on.

This is the merged-database counterpart of the per-source normalisation analysis on
the clean databases: run once on the single merged database, where cross-source
duplicates are already removed, so the four sources that carry normalisers are
**pooled** into one estimate per element (a per-source breakdown is kept alongside).
Each target is paired with the Fe and Al measured on the **same subsample and same
(bulk) fraction**, since normalisation is only valid within a comparable fraction.
Distributional outliers (see [Outlier Flagging](./outlier-flagging.qmd)) are dropped
from both the target and the normaliser. All depths are kept, reported by the EFSA
depth bands.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

ndir <- "data/merged_summary"
rd <- function(f) read.csv(file.path(ndir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
avail <- rd("merged_normalisation_availability.csv")
corr  <- rd("merged_normalisation_correlation.csv")
ratio <- rd("merged_normalisation_ratios.csv")
pairs <- rd("merged_normalisation_pairs.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
band_levels <- c("0-5", "5-40", ">40", "unknown")
src_levels  <- c("All (pooled)", "Mareano", "Vannmiljø", "ICES-DOME", "MUDAB")
theme_set(theme_minimal(base_size = 12))
```

## 1. Can the data be normalised?

How much of the bulk target data carries a co-located Fe or Al value (same
subsample), by element and depth band. Coverage is uneven: the full-suite metals
(Cobalt, Manganese, Selenium) are almost always normalisable, while Copper and Zinc,
whose bulk counts are by far the largest, carry a normaliser only about a quarter of
the time. So their normalised and un-normalised subsets should be reported apart.

```{r}
#| label: avail-table
avail |>
  mutate(Element = factor(symbol, levels = elem_levels),
         band = factor(band, levels = band_levels)) |>
  filter(band != "unknown") |>
  arrange(Element, band) |>
  select(Element, band, n_target, `% with Fe` = pct_FE, `% with Al` = pct_AL) |>
  kable(col.names = c("Element", "Depth band (cm)", "Bulk targets",
                      "% with Fe", "% with Al"),
        format.args = list(big.mark = ","))
```

## 2. How well does each normaliser work?

The r-squared of a log-log fit of metal against normaliser (bulk, at least 30
pairs). High values mean the metal is grain-size / mineralogically controlled, so
the normaliser is a sound baseline. Two things stand out in the pooled data:

- **Iron is the stronger normaliser.** For Cobalt, Copper and Zinc the Fe fit is far
  tighter than the Al fit (Co 0.70 vs 0.24, Zn 0.55 vs 0.33, Cu 0.49 vs 0.21). Al
  carries less of the grain-size signal here, so Fe is the better default carrier.
- **Molybdenum and Selenium do not normalise** on either (r-squared near zero): they
  are redox and organic controlled, not lithogenic, so Fe/Al normalisation does not
  apply to them.

```{r}
#| label: strength-table
corr |>
  filter(Source == "All (pooled)") |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  select(Element, normaliser, r2) |>
  pivot_wider(names_from = normaliser, values_from = r2) |>
  transmute(Element, `r2 (Fe)` = FE, `r2 (Al)` = AL) |>
  arrange(Element) |>
  kable()
```

```{r}
#| label: strength-fig
#| fig-cap: "Log-log metal to normaliser fit strength (r-squared), pooled and per source (bulk)."
#| fig-width: 8.5
#| fig-height: 4.5
corr |>
  filter(Source %in% c("All (pooled)", "Mareano", "ICES-DOME")) |>
  mutate(Element = factor(symbol, levels = elem_levels),
         Source = factor(Source, levels = src_levels)) |>
  ggplot(aes(Element, r2, fill = normaliser)) +
  geom_col(position = position_dodge()) +
  scale_fill_manual(values = c(AL = "#4477AA", FE = "#EE6677")) +
  facet_wrap(~ Source) +
  labs(x = NULL, y = "r-squared (log-log)", fill = "normaliser") +
  ylim(0, 1)
```

## 3. The lithogenic relationship

For the well-behaved metals, concentration scales with iron along a tight log-log
line, the natural (lithogenic) baseline. Pooled across sources and outlier-free;
points well above the line would be enrichment candidates.

```{r}
#| label: scatter-fig
#| fig-cap: "Bulk target concentration against iron, pooled across sources (log-log, linear fit)."
#| fig-width: 8
#| fig-height: 3.2
pairs |>
  filter(symbol %in% c("CO", "CU", "ZN"),
         !is.na(FE), FE > 0, value_std > 0) |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  ggplot(aes(FE, value_std)) +
  geom_point(alpha = 0.1, size = 0.5, colour = "#EE6677") +
  geom_smooth(method = "lm", formula = y ~ x, colour = "#333333", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ Element, scales = "free_y") +
  labs(x = "Fe (mg/kg, log)", y = "concentration (mg/kg, log)")
```

## 4. Normalised values (metal / normaliser)

The median metal-to-normaliser ratio (bulk, pooled), the normalised concentration on
which a local background and enrichment factor would later be built.

```{r}
#| label: ratio-table
ratio |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  select(Element, normaliser, n, median, p25, p75) |>
  pivot_wider(names_from = normaliser, values_from = c(n, median, p25, p75)) |>
  transmute(Element,
            `n (Fe)` = n_FE, `median Me/Fe` = median_FE,
            `n (Al)` = n_AL, `median Me/Al` = median_AL) |>
  arrange(Element) |>
  kable(format.args = list(big.mark = ",", scientific = TRUE))
```

::: {.callout-note}
Normalisation applies to the grain-size-controlled metals (Cobalt, Copper, Zinc, and
more weakly Manganese), and iron is the better normaliser for them here. Molybdenum
and Selenium are not lithogenically controlled, so their Fe/Al fits are near zero and
normalisation is not appropriate for them; they are kept in the tables for
completeness. The enrichment factor and pristine cutoff are built on these normalised
values at a later background stage.
:::
