---
title: "Organic Carbon"
subtitle: "Relating the target metals to total organic carbon, pooled across the merged database"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Trace metals bind to organic matter, so total organic carbon (CORG, reported as TOC
in the EFSA spec) is a third control on their concentration alongside grain size and
the Fe/Al normalisers. Where a metal tracks organic carbon but not iron or aluminium,
it is organically rather than lithogenically controlled.

This is the merged-database counterpart of the per-source organic-carbon analysis on
the clean databases: run once on the single merged database (duplicates removed), so
the four sources that carry organic carbon are **pooled** into one estimate per
element (a per-source breakdown is kept alongside). Whole-sample organic carbon is
CORG (all source TOC was harmonised to it earlier); the fines-normalised TOC63 is a
different measurand and is not used here. Each target is paired with the CORG on the
**same subsample and same (bulk) fraction**; distributional outliers (see
[Outlier Flagging](./outlier-flagging.qmd)) are dropped. CORG is converted from mg/kg
to percent. All depths are kept, reported by the EFSA depth bands.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

odir <- "data/merged_summary"
rd <- function(f) read.csv(file.path(odir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
avail <- rd("merged_organic_availability.csv")
dist  <- rd("merged_organic_distribution.csv")
corr  <- rd("merged_organic_correlation.csv")
pairs <- rd("merged_organic_pairs.csv")

elem_levels <- c("CO", "CU", "I", "MN", "MO", "SE", "ZN")
band_levels <- c("0-5", "5-40", ">40", "unknown")
src_levels  <- c("All (pooled)", "Mareano", "Vannmiljø", "ICES-DOME", "MUDAB")
theme_set(theme_minimal(base_size = 12))
```

## 1. Organic carbon coverage

How much of the bulk target data carries a co-located CORG value (same subsample), by
element and depth band. Coverage is broad: notably Copper and Zinc, normalisable on
Fe/Al only about a quarter of the time, carry organic carbon for well over half their
readings, so CORG is the practical control for them where a normaliser is missing.

```{r}
#| label: avail-table
avail |>
  mutate(Element = factor(symbol, levels = elem_levels),
         band = factor(band, levels = band_levels)) |>
  filter(band != "unknown") |>
  arrange(Element, band) |>
  select(Element, band, n_target, `% with CORG` = pct_CORG) |>
  kable(col.names = c("Element", "Depth band (cm)", "Bulk targets", "% with CORG"),
        format.args = list(big.mark = ","))
```

## 2. Organic carbon content (%)

The organic carbon percentage (bulk subsamples), pooled by depth band. Values are
typical for marine sediment (around 1 to 3 %), with organic-rich outliers up to
tens of percent.

```{r}
#| label: dist-table
dist |>
  mutate(band = factor(band, levels = band_levels)) |>
  filter(band != "unknown") |>
  arrange(band) |>
  select(band, n_subsamples, `median %` = median, p25, p75, `max %` = max) |>
  kable(col.names = c("Depth band (cm)", "Subsamples", "Median %", "P25", "P75",
                      "Max %"),
        format.args = list(big.mark = ","))
```

## 3. Which metals track organic carbon?

The r-squared of a log-log fit of metal against CORG (bulk, at least 30 pairs). The
pooled result reads directly against the [Fe/Al normalisation](./normalisation.qmd)
page:

- **Molybdenum tracks organic carbon** (r-squared 0.27) though it did **not**
  normalise on Fe or Al: it is redox / organic controlled, so CORG is its control,
  not a lithogenic normaliser.
- **Copper and Zinc** associate with organic carbon (about 0.30), comparable to their
  iron fit, and with far broader coverage, so organic carbon is their more practical
  baseline.
- **Cobalt and Manganese**, strong on iron, are weak here: lithogenic, not organic.
- **Selenium** tracks nothing pooled (grain size, Fe/Al, and now CORG all near zero),
  so it resists a simple single-covariate baseline.

```{r}
#| label: strength-table
corr |>
  filter(Source == "All (pooled)") |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  arrange(Element) |>
  transmute(Element, n, `r2 (CORG)` = r2, slope) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: strength-fig
#| fig-cap: "Log-log metal to CORG fit strength (r-squared), pooled and per source (bulk)."
#| fig-width: 8.5
#| fig-height: 4.5
corr |>
  filter(Source %in% c("All (pooled)", "Mareano", "ICES-DOME", "Vannmiljø")) |>
  mutate(Element = factor(symbol, levels = elem_levels),
         Source = factor(Source, levels = src_levels)) |>
  ggplot(aes(Element, r2)) +
  geom_col(fill = "#228833") +
  facet_wrap(~ Source) +
  labs(x = NULL, y = "r-squared (log-log)") +
  ylim(0, 1)
```

## 4. The organic relationship

For the organic-associated metals, concentration rises with organic carbon. Pooled
across sources and outlier-free, the trend is clearest for copper, zinc and
molybdenum.

```{r}
#| label: scatter-fig
#| fig-cap: "Bulk target concentration against organic carbon, pooled across sources (log y, linear fit)."
#| fig-width: 8
#| fig-height: 3.2
pairs |>
  filter(symbol %in% c("CU", "ZN", "MO"),
         !is.na(corg_pct), corg_pct > 0, value_std > 0) |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  ggplot(aes(corg_pct, value_std)) +
  geom_point(alpha = 0.08, size = 0.5, colour = "#228833") +
  geom_smooth(method = "lm", formula = y ~ x, colour = "#AA3377", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ Element, scales = "free_y") +
  labs(x = "organic carbon (%, log)", y = "concentration (mg/kg, log)")
```

::: {.callout-note}
Read together with the [Fe/Al Normalisation](./normalisation.qmd) page, the three
controls sort the metals: Cobalt and Manganese are lithogenic (iron), Molybdenum is
organic (CORG), and Copper and Zinc respond to both. This split guides which control
to use per element when defining a local background. Iodine and Selenium remain sparse
or unresponsive and need separate treatment.
:::
