---
title: "DB Schema (Refined)"
subtitle: "The analysis-ready mart derived from the merged database"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
.mermaid { max-width: 680px; margin-left: auto; margin-right: auto; }
.mermaid svg { max-width: 100% !important; width: 100%; height: auto; }
```

The **refined** database (`multised_refined.sqlite`) is a restructured, slimmed copy of
the [merged database](./db-schema-merged.qmd), built for the later work on identifying
pristine / background sediments. It bakes in the reusable results of the merged-database
analyses so those are not recomputed, and drops the data that work did not use.

## The rule behind the refactor

One rule decides what the refined database holds:

> **Bake in facts and already-computed reusable values; defer every classification or
> threshold.**

So it carries the **normalisation ratios** (metal / Fe, metal / Al, metal / organic
carbon), the grain-size fines, the aquaculture link and the repeat-sampling grouping,
all as ready-to-use columns. It carries **none** of the decisions the pristine work will
own: no depth bands, no offshore or pressure flags, no enrichment-factor thresholds, no
pristine label, and no cluster labels (those re-run cheaply from the ratios). The raw
material for all of those (ratios, distances, raw depth) is present; the calls are not
made here.

## What changed from the merged database

- **`measurement` splits into `bulk_measurement` and `sieved_measurement`.** The two
  fractions are never pooled, so the split makes that structural; the sieve columns live
  only on the sieved table. Both hold **only the 7 target elements**.
- **Fe, Al and organic carbon leave `measurement`.** They were only ever normalisers, so
  they become a compact **`normaliser`** table (one wide row per subsample and fraction),
  and each target row gains the baked **`ratio_fe` / `ratio_al` / `ratio_corg`**.
- **`grain_size_fraction` is dropped.** Every analysis used the derived `fines_lt63` (kept
  on `subsample`), never the raw curve.
- **`site` gains `aqua_id`** (a link to the new **`aquaculture`** table) and
  **`repeat_group` / `n_years`** (the repeat-sampled-location grouping).
- **`element` is trimmed to the 7 targets**, and `method` to the methods those targets use.

## Schema diagram

```{mermaid}
erDiagram
  dataset     ||--o{ event              : "dataset_id"
  site        ||--o{ event              : "site_id"
  aquaculture ||--o{ site               : "aqua_id"
  event       ||--o{ subsample          : "event_id"
  subsample   ||--o{ bulk_measurement   : "subsample_id"
  subsample   ||--o{ sieved_measurement : "subsample_id"
  subsample   ||--o{ normaliser         : "subsample_id"
  element     ||--o{ bulk_measurement   : "symbol"
  element     ||--o{ sieved_measurement : "symbol"
  element     ||--o{ method             : "symbol"
  method      ||--o{ bulk_measurement   : "method_id"
  method      ||--o{ sieved_measurement : "method_id"

  aquaculture {
    INTEGER aqua_id PK
  }
  site {
    INTEGER site_id PK
    INTEGER aqua_id FK
    TEXT repeat_group
  }
  event {
    INTEGER event_id PK
    INTEGER dataset_id FK
    INTEGER site_id FK
  }
  subsample {
    INTEGER subsample_id PK
    INTEGER event_id FK
    REAL fines_lt63
  }
  normaliser {
    INTEGER subsample_id FK
    TEXT frac_class
    REAL fe
    REAL al
    REAL corg
  }
  bulk_measurement {
    INTEGER measurement_id PK
    INTEGER subsample_id FK
    TEXT symbol FK
    REAL ratio_fe
  }
  sieved_measurement {
    INTEGER measurement_id PK
    INTEGER subsample_id FK
    TEXT symbol FK
    REAL sieve_um_std
  }
  element {
    TEXT symbol PK
  }
  method {
    INTEGER method_id PK
    TEXT symbol FK
  }
```

```{r}
library(tibble)
library(knitr)
```

## bulk_measurement / sieved_measurement

One row per target-element reading, split by fraction. Both carry the baked
normalisation ratios; `sieved_measurement` adds the sieve cutoff columns.

```{r}
tribble(
  ~Column,             ~Type,     ~Meaning,
  "measurement_id",    "INTEGER", "Primary key (from the merged database).",
  "subsample_id",      "INTEGER", "-> subsample.",
  "symbol",            "TEXT",    "Target element (-> element); one of CO CU I MN MO SE ZN.",
  "value",             "REAL",    "Value in the original unit.",
  "unit",              "TEXT",    "Original reporting unit.",
  "value_std",         "REAL",    "Standardised value, mg/kg.",
  "unit_std",          "TEXT",    "Always mg/kg.",
  "value_sd",          "REAL",    "SD across replicates, else NULL.",
  "n_rep",             "INTEGER", "Replicates averaged (1 = single).",
  "value_uncrt",       "REAL",    "Analytical uncertainty, mg/kg, else NULL.",
  "matrix",            "TEXT",    "Raw ICES fraction code (provenance).",
  "method_id",         "INTEGER", "-> method.",
  "ratio_fe",          "REAL",    "value_std / Fe (same subsample, fraction), else NULL.",
  "ratio_al",          "REAL",    "value_std / Al, else NULL.",
  "ratio_corg",        "REAL",    "value_std / organic carbon, else NULL.",
  "source",            "TEXT",    "Source database name.",
  "src_measurement_id","TEXT",    "Original key (provenance).",
  "outlier_flag",      "TEXT",    "Distributional outlier marker: 'high' / 'low', else NULL.",
  "sieve_um_std",      "REAL",    "sieved_measurement only: harmonised sieve cutoff (um).",
  "sieve_class",       "TEXT",    "sieved_measurement only: sieve label, e.g. '<63um'."
) |> kable()
```

## normaliser

The Fe / Al / organic-carbon values, reshaped from the merged `measurement` rows to one
wide row per **subsample and fraction** (the grain a ratio needs, since a subsample can
carry both fractions). Multi-method values are mean-collapsed.

```{r}
tribble(
  ~Column,        ~Type,     ~Meaning,
  "subsample_id", "INTEGER", "-> subsample.",
  "frac_class",   "TEXT",    "'bulk' or 'sieved'.",
  "fe",           "REAL",    "Iron, mg/kg, else NULL.",
  "al",           "REAL",    "Aluminium, mg/kg, else NULL.",
  "corg",         "REAL",    "Organic carbon, mg/kg, else NULL."
) |> kable()
```

## site (additions)

The merged `site` columns, plus:

```{r}
tribble(
  ~Column,        ~Type,     ~Meaning,
  "aqua_id",      "INTEGER", "-> aquaculture: the nearest aquaculture site (Norway only, else NULL).",
  "repeat_group", "TEXT",    "Location cell (lat/lon to 2 dp, ~1.1 km) for repeat-sampling grouping.",
  "n_years",      "INTEGER", "Distinct sampling years at that cell (over all dated events)."
) |> kable()
```

## aquaculture

The Norwegian aquaculture reference, imported so `site.aqua_id` can point to it.

```{r}
tribble(
  ~Column,           ~Type,     ~Meaning,
  "aqua_id",         "INTEGER", "Primary key.",
  "loknr",           "TEXT",    "Site (locality) number.",
  "name",            "TEXT",    "Site name.",
  "latitude",        "REAL",    "Latitude, decimal degrees.",
  "longitude",       "REAL",    "Longitude, decimal degrees.",
  "start_year",      "INTEGER", "First year of operation, where known.",
  "end_year",        "INTEGER", "Last year of operation, where known.",
  "active",          "INTEGER", "1 = currently active, 0 = closed.",
  "capacity_tonnes", "REAL",    "Licensed capacity, tonnes.",
  "fish_types",      "TEXT",    "Species farmed.",
  "placement",       "TEXT",    "Placement descriptor (e.g. sea, land).",
  "county",          "TEXT",    "County.",
  "municipality",    "TEXT",    "Municipality."
) |> kable()
```

## Carried over

`subsample` keeps its physical columns (`depth_from` / `depth_to`, the exist flags,
`fines_lt63` / `fines_basis`); `dataset` and `event` are unchanged; `method` keeps the
detection limits (`lod` / `loq`). Their columns are as on the
[merged schema](./db-schema-merged.qmd). Depths are kept **raw** (no bands), and no Fe /
Al / organic values are duplicated onto the target rows: they live once in `normaliser`,
reachable by the subsample and fraction.

::: {.callout-note}
The refined database is deliberately incomplete by design: it stops at the reusable
building blocks. The pristine / background analyses that consume it (enrichment factors,
depth bands, pressure and pristine classification) live on a separate site, so the choices
they make are never frozen into the data here. How the refined database is assembled step
by step is on [Building the Refined DB](./refined-build.qmd).
:::
