---
title: "Building the Refined DB"
subtitle: "How the merged database is restructured into the refined mart, and what is kept"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The [refined database](./refined-schema.qmd) is built from the merged database in a short
sequence of steps: restructure the tables, reshape the normalisers, bake the ratios, link
the aquaculture reference, group the repeat-sampled locations, and reconcile. This page
reports what each step keeps and drops, so every merged row is accounted for.

```{r}
#| label: setup
library(dplyr)
library(ggplot2)
library(knitr)

fdir <- "data/merged_summary"
rd <- function(f) read.csv(file.path(fdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
tbl   <- rd("refined_tables.csv")
recon <- rd("refined_reconciliation.csv")
cov   <- rd("refined_coverage.csv")
theme_set(theme_minimal(base_size = 12))
```

## What happens to each table

```{r}
#| label: tables
tbl |>
  transmute(Table = table,
            `merged rows` = ifelse(is.na(merged_rows), "", format(merged_rows, big.mark = ",")),
            `refined rows` = ifelse(is.na(refined_rows), "", format(refined_rows, big.mark = ",")),
            Disposition = disposition) |>
  kable(align = "lllr")
```

The seven target elements stay in a single `measurement` table (bulk and sieved told
apart by `frac_class`); the Fe / Al / organic-carbon rows become the `normaliser`;
`grain_size_fraction` is dropped in favour of the derived fines already on `subsample`;
and the dimensions are carried, with `element` and `method` trimmed to the targets.

## Every merged measurement row accounted for

The merged `measurement` table has one row per reading. Each goes to exactly one place, or
is dropped for a stated reason:

```{r}
#| label: reconciliation
recon |>
  transmute(`Merged rows` = group, n = rows, Destination = destination) |>
  kable(format.args = list(big.mark = ","))
```

```{r}
#| label: recon-check
total <- sum(recon$rows)
```

That is **`r format(total, big.mark = ",")`** rows in total, the full merged measurement
table. The only rows dropped are `TOC63` (organic carbon reported on the sieved fraction,
a different quantity from whole-sample organic carbon and not used as a normaliser).

## Coverage of the baked values

The ratios and normalisers are only as complete as the underlying data: a target row gets
a ratio only where the matching normaliser was measured in the same subsample and fraction.
The chart shows how much of each is populated.

```{r}
#| label: coverage-fig
#| fig-cap: "Percentage populated for each baked field, by table. Organic carbon is the best-covered normaliser for bulk, aluminium for sieved; the aquaculture link covers the Norway sites."
#| fig-width: 8
#| fig-height: 4.2
cov |>
  mutate(lab = paste(scope, field),
         lab = reorder(lab, pct)) |>
  ggplot(aes(pct, lab)) +
  geom_col(fill = "#4477AA") +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.15, size = 3) +
  scale_x_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.12))) +
  labs(x = "% populated", y = NULL)
```

```{r}
#| label: coverage-table
cov |>
  transmute(Scope = scope, Field = field,
            populated = n_present, of = n_total, `%` = pct) |>
  kable(format.args = list(big.mark = ","))
```

::: {.callout-note}
Coverage is a property of the source data, not of the build: a missing ratio means the
normaliser was never measured for that sample, not that a step failed. The values that do
exist are carried once, so the pristine analyses filter and aggregate them without
recomputing. The schema itself is on [DB Schema (Refined)](./refined-schema.qmd).
:::
