---
title: "Repeat-Sampled Sites"
subtitle: "Trends over time at places sampled in more than one year, with location held roughly fixed"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The [Sampling Year](./temporal.qmd) page pools every bulk measurement against its
year, so "year" is tangled up with **where** each source sampled in each decade
(one source owns the long Norwegian series, another only a single recent window).
It can say there is no strong secular drift once outliers are removed, but not
whether inputs changed **at a given place**, because the places themselves change
from year to year.

This page fixes that. It keeps only locations that were sampled in **more than one
year** and measures the trend **inside** each location, so the differences between
places are differenced out and what remains is closer to a real change over time at
a fixed spot.

A "location" here is a grid cell: latitude and longitude rounded to two decimals
(about 1.1 km), coarser than the exact site key so nearby revisits of the same
station group together. A cell is used for an element when it spans at least three
distinct years and holds at least eight bulk measurements of that element.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

sdir <- "data/merged_summary"
rd <- function(f) read.csv(file.path(sdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
trends <- rd("merged_siteyears_trends.csv")
cells  <- rd("merged_siteyears_cells.csv")
pooled <- rd("merged_siteyears_pooled.csv")
meta   <- rd("merged_siteyears_meta.csv")

elem_levels <- c("CO", "CU", "MN", "ZN")
BLUE <- "#4477AA"; RED <- "#AA3377"
theme_set(theme_minimal(base_size = 12))
```

At this resolution the merged database holds
**`r format(meta$cells_multiyear, big.mark = ",")`** locations sampled in three or
more years (out of `r format(meta$cells_total, big.mark = ",")` cells in all).
Copper and zinc carry the analysis; cobalt and manganese have enough repeat
sampling to report with caution; molybdenum, selenium and iodine are barely
revisited and drop out.

## A. Do raw levels trend at a fixed location?

For each cell and element we take the Spearman correlation of the value with year,
then look at the **distribution of those within-cell correlations** across cells.
A median near zero means that, once you stand in one place, the metal is neither
systematically rising nor falling.

```{r}
#| label: a-table
trends |>
  filter(measure == "value") |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  filter(!is.na(Element)) |>
  transmute(Element, `cells` = n_cells, `median rho` = median_rho,
            `% declining` = round(100 * frac_declining),
            `% rising` = round(100 * frac_rising),
            `median span (yr)` = median_span, `sign-test p` = p_sign) |>
  arrange(Element) |>
  kable(caption = "Within-cell trend of the raw value vs year, summarised across cells.")
```

Copper and zinc, the well-covered pair, sit almost exactly at zero (about half the
cells rise, half fall): **the raw levels are flat over time at repeat-sampled
places.** Manganese drifts gently down (a modest majority of cells declining);
cobalt looks to rise but rests on only a handful of cells, so read it lightly.

```{r}
#| label: a-fig
#| fig-cap: "Distribution of within-cell value-vs-year correlations, one point per location. The box is the interquartile range; the dashed line is no trend. A box centred on zero means levels are flat once location is fixed."
#| fig-width: 7.5
#| fig-height: 3.4
cells |>
  filter(measure == "value", symbol %in% elem_levels) |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  ggplot(aes(rho, Element)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_boxplot(outlier.size = 0.6, fill = "grey92", width = 0.55) +
  labs(x = "within-cell Spearman rho (value vs year)", y = NULL) +
  xlim(-1, 1)
```

## B. Does a trend appear after normalisation?

The same within-cell correlation, now on the **metal / iron ratio** (iron measured
in the same subsample, the best grain-size carrier from the
[Fe/Al Normalisation](./normalisation.qmd) page). Normalising strips out how much
fine, metal-holding sediment happened to be in each grab, so a trend the raw value
hides can surface.

```{r}
#| label: b-table
trends |>
  filter(measure == "enrich_FE") |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  filter(!is.na(Element)) |>
  transmute(Element, `cells` = n_cells, `median rho` = median_rho,
            `% declining` = round(100 * frac_declining),
            `% rising` = round(100 * frac_rising),
            `sign-test p` = p_sign) |>
  arrange(Element) |>
  kable(caption = "Within-cell trend of the metal / Fe enrichment ratio vs year.")
```

Zinc is the clear result: **flat as a raw value, but Zn/Fe falls within the same
locations** (a median correlation near -0.33, roughly seven cells in ten declining,
sign-test p well below 0.01). Standing in one place, the zinc load relative to its
lithogenic carrier has come down over the years, the improvement signal the pooled
[Sampling Year](./temporal.qmd) page could only hint at. Copper leans the same way
more weakly; manganese and cobalt are indistinct on the fewer cells that carry iron.

```{r}
#| label: b-fig
#| fig-cap: "Within-cell metal/Fe trend by element. Zinc sits clearly left of zero (declining enrichment at fixed locations); the others straddle it."
#| fig-width: 7.5
#| fig-height: 3.4
cells |>
  filter(measure == "enrich_FE", symbol %in% elem_levels) |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  ggplot(aes(rho, Element)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_boxplot(outlier.size = 0.6, fill = "grey92", width = 0.55) +
  labs(x = "within-cell Spearman rho (metal / Fe vs year)", y = NULL) +
  xlim(-1, 1)
```

## C. The shared drift, with location removed

To combine the locations without letting the busiest cell dominate, we express each
measurement as a log10 deviation from **its own cell's median**, then take the
yearly median of that deviation across all cells. This removes the mean level of
each place and leaves only the drift they share. A value of 0.1 on this scale is
about a 25 percent departure from the local baseline.

```{r}
#| label: c-fig
#| fig-cap: "Yearly median location-anomaly (log10 deviation from each location's own median), copper and zinc. The band near zero shows how little the raw level moves at fixed places; the loess line traces the gentle drift."
#| fig-width: 8
#| fig-height: 3.6
pooled |>
  filter(symbol %in% c("CU", "ZN")) |>
  mutate(Element = factor(symbol, levels = c("CU", "ZN"))) |>
  ggplot(aes(year, median_anom)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_point(aes(size = n_cells), alpha = 0.45, colour = BLUE) +
  geom_smooth(method = "loess", formula = y ~ x, colour = RED, se = FALSE,
              linewidth = 0.7) +
  scale_size_area(max_size = 3, guide = "none") +
  facet_wrap(~ Element) +
  labs(x = "sampling year", y = "median log10 anomaly from local baseline")
```

Around nine years in ten sit within plus or minus 0.1 log10 (roughly 25 percent) of
the local baseline, and the loess line is nearly flat: raw copper and zinc levels
barely move at repeat-sampled places. The few larger excursions are sparse early
years built on only a couple of locations (small points). The real temporal story is
not in the raw concentration but in the **Fe-normalised zinc** of section B.

::: {.callout-note}
Holding location fixed is a big improvement on the pooled page, but not a controlled
experiment: a cell can still contain a few sub-stations, and differencing location
does not difference out changes in gear or laboratory within a cell over time. These
remain associations. The one robust, direction-clear result is the decline in
zinc relative to iron at repeat-sampled locations; everything else is flat or too
thinly revisited to call. Molybdenum, selenium and iodine have too few repeat-sampled
locations to appear.
:::
