---
title: "Spatial Hotspots"
subtitle: "Where elevated readings cluster geographically, element by element"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The [Geochemical Facies](./geochemical-facies.qmd) page clusters samples in chemistry
space. This one clusters them in **geographic** space, one element at a time, to
locate the **hotspots**: regions where elevated readings pile up rather than appearing
as one-off highs.

The recipe is simple. For each element and fraction, keep only the **elevated**
readings (the top fifth of the metric), project their positions to kilometres, and run
DBSCAN, a density clustering method. A dense group of nearby elevated readings becomes
a hotspot; an elevated reading with no elevated neighbours is left as "isolated" (a
one-off, not a hotspot). The metric follows the same rule as the facies page: **bulk**
on the iron ratio (metal / Fe, so a hotspot is an *enrichment* hotspot with grain size
removed), **sieved** on the **raw** value. Bulk and sieved are treated separately, and
distributional outliers are dropped.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

hdir <- "data/merged_summary"
rd <- function(f) read.csv(file.path(hdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
summ <- rd("merged_hotspots_summary.csv")
toph <- rd("merged_hotspots_top.csv")
pts  <- rd("merged_hotspots_points.csv")
meta <- rd("merged_hotspots_meta.csv")

theme_set(theme_minimal(base_size = 12))
```

Elevated is the **top `r round(100 * (1 - meta$high_quantile))` percent** of the
metric; DBSCAN uses a `r meta$eps_km` km neighbourhood and a minimum of
`r meta$minpts` points to seed a hotspot.

## How much clusters

Across every element, most elevated readings do fall into a hotspot rather than
standing alone. That is partly the phenomenon and partly the sampling: the survey
coverage is itself spatially clustered (dense fjords and monitoring transects, sparse
open sea), so at a fixed radius a large share of any subset groups up. The useful
output is therefore not the clustered fraction but **which** hotspots are largest and
where they sit.

```{r}
#| label: summary-table
summ |>
  transmute(Element = symbol, Fraction = frac_class, `elevated n` = n_high,
            hotspots = n_hotspots, `% in a hotspot` = pct_clustered) |>
  arrange(Fraction, Element) |>
  kable(format.args = list(big.mark = ","))
```

## The largest hotspots

Each hotspot is named by the sea area most of its samples fall in. Copper and zinc,
the two contamination-relevant metals, are shown here (manganese and cobalt form
their own hotspots, counted in the table above). The median value is the raw
concentration of the members, so the
enrichment-based bulk hotspots still report in mg/kg.

```{r}
#| label: top-table
toph |>
  filter(symbol %in% c("CU", "ZN"), rank <= 5) |>
  transmute(Element = symbol, Fraction = frac_class, Rank = rank, n,
            `sea area` = sea_area, `median mg/kg` = median_value) |>
  kable(caption = "Five largest copper and zinc hotspots per fraction.")
```

The bulk (enrichment) hotspots spread across the shelf, the North Sea, the Norwegian
Sea and the North Atlantic margin, wherever metal runs high relative to iron. The
sieved hotspots, by contrast, sit almost entirely in the **North Sea**: that is where
the sieved-fraction data is concentrated (the German and OSPAR monitoring), so the
sieved map reflects where that programme sampled as much as where the metal is.

## Maps

Elevated readings for copper and zinc, coloured by whether DBSCAN placed them in a
hotspot (red) or left them isolated (grey). Rows are the fraction, columns the element.

```{r}
#| label: hotspot-map
#| fig-cap: "Elevated copper and zinc readings: red points belong to a spatial hotspot, grey points are isolated highs. The sieved row is dominated by the North Sea monitoring footprint."
#| fig-width: 8.5
#| fig-height: 8
d <- pts |>
  filter(symbol %in% c("CU", "ZN")) |>
  mutate(kind = ifelse(hotspot_id > 0, "hotspot", "isolated")) |>
  arrange(kind)   # draw hotspot points on top

world <- ggplot2::map_data("world")

ggplot() +
  geom_polygon(data = world, aes(long, lat, group = group),
               fill = "grey93", colour = "grey80", linewidth = 0.2,
               inherit.aes = FALSE) +
  geom_point(data = d, aes(longitude, latitude, colour = kind),
             size = 0.7, alpha = 0.7) +
  coord_quickmap(xlim = c(-20, 40), ylim = c(36, 82)) +
  scale_colour_manual(values = c(hotspot = "#EE6677", isolated = "grey55"),
                      name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  facet_grid(frac_class ~ symbol) +
  labs(x = NULL, y = NULL) +
  theme(panel.grid = element_line(colour = "grey93"))
```

::: {.callout-note}
DBSCAN uses one global neighbourhood radius, but marine sampling density is very
uneven, so a fixed radius over-merges where sampling is dense and can miss a real but
thinly sampled hotspot; the largest, named hotspots are robust to this, the smallest
are not. A variable-density alternative (HDBSCAN) was checked as a cross-method test:
it agreed with DBSCAN on the uniform sieved data, and on the uneven bulk data it
sharpened some labels (surfacing enclosed-sea zinc hotspots in the Skagerrak and the
Baltic) but was sensitive to its cluster-size setting and gave no decisive
improvement, so DBSCAN is kept. A bulk hotspot is an **enrichment** hotspot (metal
relative to iron), not
necessarily a high total-concentration one; the sieved hotspots track the North Sea
monitoring footprint as much as the chemistry. Interpolating these points into a
continuous surface is deliberately avoided: a marine interpolation would smear values
across peninsulas and between disconnected basins. Molybdenum, selenium and iodine are
too sparsely measured to support spatial density clustering.
:::
