---
title: "Depth and Coast"
subtitle: "How the geographic setting relates to enrichment and the normalisers"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Do the geographic attributes **depth** and **distance to coast** leave a gradient
after normalisation? Two questions, on bulk samples, pooled across the merged database
(distributional outliers dropped):

- **(A, the point)** does the **enrichment** (metal / normaliser) change offshore and
  with depth? A ratio that falls offshore is the coastal-pollution gradient a pristine
  background rests on: cleaner offshore sediment. A flat ratio means normalisation
  already absorbed the spatial structure, so the metal was lithogenic, not enriched.
- **(B, a check)** do the **normalisers and covariates** themselves (Al, Fe, organic
  carbon, fines) vary with depth and distance to coast? They do, via grain size,
  which is why the raw metal does, and why (A) works on the ratio.

Iron is the primary normaliser here (it carried the grain-size signal best in the
[Fe/Al Normalisation](./normalisation.qmd) analysis); aluminium is shown alongside
because the choice turns out to matter spatially.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

sdir <- "data/merged_summary"
rd <- function(f) read.csv(file.path(sdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
enr <- rd("merged_spatial_enrichment.csv")
cov <- rd("merged_spatial_covariate.csv")
prs <- rd("merged_spatial_pairs.csv")

elem_levels <- c("CO", "CU", "MN", "MO", "SE", "ZN")
theme_set(theme_minimal(base_size = 12))
```

## A. Enrichment versus depth and distance to coast

Spearman correlation of the metal / normaliser ratio with distance to coast and with
depth (bulk, pooled, at least 50 per group). **Negative means the ratio falls offshore
/ with depth: cleaner offshore.**

```{r}
#| label: a-table
enr |>
  filter(Source == "All (pooled)", normaliser == "Fe") |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  select(Element, variable, rho) |>
  pivot_wider(names_from = variable, values_from = rho) |>
  transmute(Element, `rho vs coast` = dist_to_coast, `rho vs depth` = depth) |>
  arrange(Element) |>
  kable(caption = "Metal / Fe enrichment vs geography (pooled Spearman rho).")
```

Iron-normalised, **zinc shows a steep coastal gradient** (its enrichment falls sharply
offshore), copper and manganese a milder one, while **cobalt is flat**: cobalt is
purely lithogenic, so Fe-normalisation leaves no spatial residual, exactly the null
case. Molybdenum and selenium also fall offshore, but through their organic / redox
affinity (organic carbon is higher near the coast) rather than metal pollution.

```{r}
#| label: a-scatter
#| fig-cap: "Iron-normalised enrichment against distance to coast, pooled (log y, linear fit). Zinc and copper fall offshore; cobalt is flat."
#| fig-width: 8
#| fig-height: 3
prs |>
  filter(symbol %in% c("ZN", "CU", "CO"),
         !is.na(ratio_FE), ratio_FE > 0, !is.na(dist_to_coast)) |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  ggplot(aes(dist_to_coast, ratio_FE)) +
  geom_point(alpha = 0.06, size = 0.4, colour = "#4477AA") +
  geom_smooth(method = "lm", formula = y ~ x, colour = "#AA3377", se = FALSE) +
  scale_y_log10() +
  facet_wrap(~ Element, scales = "free_y") +
  labs(x = "distance to coast (km)", y = "metal / Fe (log)")
```

### The normaliser choice matters spatially

The same zinc enrichment looks completely different depending on the normaliser: on
iron the offshore gradient is strong, but on aluminium it nearly vanishes. That is
because aluminium itself declines offshore (see B below), so dividing by it cancels
the coastal signal, whereas iron is roughly flat against the coast and preserves it.
Iron is therefore the better normaliser for reading the coastal-pollution gradient.

```{r}
#| label: a-fe-vs-al
enr |>
  filter(Source == "All (pooled)", variable == "dist_to_coast",
         symbol %in% c("CU", "ZN", "CO")) |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  select(Element, normaliser, rho) |>
  pivot_wider(names_from = normaliser, values_from = rho) |>
  transmute(Element, `rho vs coast (Fe-norm)` = Fe, `rho vs coast (Al-norm)` = Al) |>
  arrange(Element) |>
  kable()
```

## B. Do the normalisers themselves vary with depth and coast?

Spearman correlation of aluminium, iron, organic carbon and fines with depth and
distance to coast (pooled). There is clear spatial structure: fines and iron increase
into the deeper, muddier offshore basins, while aluminium and organic carbon decline
offshore. This confirms why the raw metals carry a spatial signal, and why aluminium,
declining offshore, masks the coastal enrichment that iron reveals.

```{r}
#| label: b-table
cov |>
  select(factor, variable, rho) |>
  pivot_wider(names_from = variable, values_from = rho) |>
  transmute(Factor = factor, `rho vs coast` = dist_to_coast, `rho vs depth` = depth) |>
  arrange(Factor) |>
  kable()
```

::: {.callout-note}
Depth and distance to coast are strongly collinear (offshore is deeper) and both
co-vary with grain size, so these are associations, not isolated effects; separating
them needs a multivariate model, a later step. Depth is modelled (GEBCO) and absent on
a few land-rounded sites. The offshore decline in iron-normalised enrichment (clearest
for zinc and copper) means the most enriched sediment sits near the coast and the
offshore floor is the natural background candidate.
:::
