---
title: "Sampling Year"
subtitle: "How enrichment and the normalisers relate to when the sample was taken"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Does sampling year leave a trend after normalisation, the time analogue of the
[Depth and Coast](./spatial.qmd) page? Two questions, on bulk samples, pooled across
the merged database (distributional outliers dropped):

- **(A, the point)** does the **enrichment** (metal / normaliser) trend over the
  years? A ratio that falls with year would be a real improvement signal, declining
  input relative to the lithogenic carrier; a flat ratio means there is no temporal
  drift to see.
- **(B, a check)** do the **normalisers and covariates** themselves (Al, Fe, organic
  carbon, fines) trend with year? A drift here would mean the sampling or method mix
  changed over time and would confound (A).

Iron is the primary normaliser (best grain-size carrier in the
[Fe/Al Normalisation](./normalisation.qmd) analysis). **This is an unbalanced
observational series, not a fixed monitoring station**: the years 1975 to 2025 mix
different sources, regions and methods, and which source dominates shifts by decade,
so any trend is an association, not a station record.

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

tdir <- "data/merged_summary"
rd <- function(f) read.csv(file.path(tdir, f), stringsAsFactors = FALSE,
                           check.names = FALSE)
enr  <- rd("merged_temporal_enrichment.csv")
cov  <- rd("merged_temporal_covariate.csv")
yr   <- rd("merged_temporal_yearly.csv")

elem_levels <- c("CO", "CU", "MN", "MO", "SE", "ZN")
theme_set(theme_minimal(base_size = 12))
```

## A. Enrichment versus sampling year

Spearman correlation of the metal / iron ratio with year (bulk, pooled, at least 50
per group). **Negative means enrichment declines over time.** The trends are **weak**:
most sit within roughly plus or minus 0.2. Cobalt, copper, zinc and manganese drift
gently down, consistent with slowly falling inputs; molybdenum is flat to rising
(redox, not pollution). Selenium shows the steepest decline, but over a short, thin
run of years (see the figure), so it is the least robust of the set.

```{r}
#| label: a-table
enr |>
  filter(Source == "All (pooled)") |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  select(Element, normaliser, rho) |>
  pivot_wider(names_from = normaliser, values_from = rho) |>
  transmute(Element, `rho vs year (Fe-norm)` = Fe, `rho vs year (Al-norm)` = Al) |>
  arrange(Element) |>
  kable(caption = "Metal / normaliser enrichment vs year (pooled Spearman rho).")
```

The iron and aluminium columns disagree in sign for some metals (zinc falls on iron
but rises on aluminium). That is not a real reversal but the unbalanced series showing
through: the iron-normalisable and aluminium-normalisable samples are different
populations, drawn from different sources and decades, so the two normalisers see
different slices of time. It is another reason to read these as weak associations.

```{r}
#| label: a-fig
#| fig-cap: "Yearly median iron-normalised enrichment by element (log y, loess smooth). Point size scales with the number of samples that year."
#| fig-width: 8.5
#| fig-height: 5
yr |>
  mutate(Element = factor(symbol, levels = elem_levels)) |>
  ggplot(aes(year, median_ratio_FE)) +
  geom_point(aes(size = n), alpha = 0.4, colour = "#4477AA") +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#AA3377", se = FALSE,
              linewidth = 0.7) +
  scale_y_log10() +
  scale_size_area(max_size = 3, guide = "none") +
  facet_wrap(~ Element, scales = "free_y") +
  labs(x = "sampling year", y = "median metal / Fe (log)")
```

## B. Do the normalisers themselves drift with year?

Spearman correlation of aluminium, iron, organic carbon and fines with year (pooled).
Reassuringly, all four are close to flat (only fines drifts slightly, as sampling
leaned marginally toward muddier sites over time). Because the covariates do not drift,
the weak enrichment trends in A are not a normaliser-drift artefact; they are what
little temporal signal the pooled data carries.

```{r}
#| label: b-table
cov |>
  transmute(Factor = factor, `rho vs year` = rho, n) |>
  arrange(Factor) |>
  kable(format.args = list(big.mark = ","))
```

::: {.callout-note}
The temporal signal is weak and the series is unbalanced, so this page is best read as
context rather than a trend record: it says there is no strong secular drift in the
normalised metals once outliers are removed, not that inputs did or did not change at
any given place. A real trend analysis would fix the station and method, a monitoring
design this merged, opportunistic dataset does not provide. Iodine has too few
year-tagged bulk samples to report.
:::
